---
title: 'Modeling Chronic Toxicity: A comparison of experimental variability with (Q)SAR/read-across predictions'
author: 
    - Christoph Helma^1^
    - David Vorgrimmler^1^
    - Denis Gebele^1^
    - Martin Gütlein^2^
    - Barbara Engeli^3^
    - Jürg Zarn^3^
    - Benoit Schilter^4^
    - Elena Lo Piparo^4^
include-before: ^1^ in silico toxicology gmbh,  Basel, Switzerland\newline^2^ Inst. f. Computer Science, Johannes Gutenberg Universität Mainz, Germany\newline^3^ Federal Food Safety and Veterinary Office (FSVO) , Risk Assessment Division , Bern , Switzerland\newline^4^ Chemical Food Safety Group, Nestlé Research Center, Lausanne, Switzerland
keywords: (Q)SAR, read-across, LOAEL, experimental variability
date: \today
abstract: |
  This study compares the accuracy of (Q)SAR/read-across predictions with the
  experimental variability of chronic lowest-observed-adverse-effect levels
  (LOAELs) from *in vivo* experiments. We could demonstrate that predictions of
  the lazy structure-activity relationships (`lazar`) algorithm within the
  applicability domain of the training data have the same variability as the
  experimental training data. Predictions with a lower similarity threshold
  (i.e. a larger distance from the applicability domain) are also significantly
  better than random guessing, but the errors to be expected are higher and
  a manual inspection of prediction results is highly recommended.

documentclass: achemso
bibliography: references.bibtex
figPrefix: Figure
eqnPrefix: Equation
tblPrefix: Table
colorlinks: true
output:
  pdf_document:
    fig_caption: yes
header-includes:
  - \usepackage{a4wide}
  - \linespread{2}
  - \usepackage{lineno}
  - \linenumbers
...

```{r echo=F}
rsquare <- function(x,y) { cor(x,y,use='complete')^2 }
rmse <-  function(x,y) { sqrt(mean((x-y)^2,na.rm=TRUE)) }

m = read.csv("data/mazzatorta_log10.csv",header=T)
s = read.csv("data/swiss_log10.csv",header=T)
t = read.csv("data/test_log10.csv",header=T)
c = read.csv("data/training_log10.csv",header=T)
```

Introduction
============

Relying on standard animal toxicological testing for chemical hazard
identification and characterization is increasingly questioned on both
scientific and ethical grounds. In addition, it appears obvious that from
a resource perspective, the capacity of standard toxicology to address the
safety of thousands of untested chemicals [@Fowler2011] to which human may be
exposed is very limited. It has also been recognized that getting rapid insight
on toxicity of chemicals in case of emergency safety incidents or for early
prioritization in research and development (safety by design) is a big
challenge mainly because of the time and cost constraints associated with the
generation of relevant animal data. In this context, alternative approaches to
obtain timely and fit-for-purpose toxicological information are being
developed. Amongst others *in silico*
toxicology methods are considered highly
promising. Importantly, they are raising more and more interests
and getting increased acceptance in various regulatory (e.g.
[@ECHA2008, @EFSA2016, @EFSA2014, @HealthCanada2016, @OECD2015]) and industrial (e.g.
[@Stanton2016, @LoPiparo2011]) frameworks.

For a long time already, computational methods have been an integral
part of pharmaceutical discovery pipelines, while in chemical food
safety their actual potentials emerged only recently [@LoPiparo2011].
In this field, an application considered critical is in the
establishment of levels of safety concern in order to rapidly and
efficiently manage toxicologically uncharacterized chemicals identified
in food. This requires a risk-based approach to benchmark exposure with
a quantitative value of toxicity relevant for risk assessment [@Schilter2014].
Since chronic studies have the highest power (more animals per group and more endpoints than other studies) and because
long-term toxicity studies are often the most sensitive in food
toxicology databases, predicting chronic toxicity is of prime
importance. Up to now, read-across and Quantitative Structure Activity
Relationships (QSAR) have been the most used *in silico* approaches to
obtain quantitative predictions of chronic toxicity.

The quality and reproducibility of (Q)SAR and read-across predictions
has been a continuous and controversial topic in the toxicological
risk-assessment community. Although model predictions can be validated
with various procedures, to review results in context of experimental
variability has actually been rarely done or attempted. With missing
information about the variability of experimental toxicity data it is
hard to judge the performance of predictive models objectively and it is
tempting for model developers to use aggressive model optimisation
methods that lead to impressive validation results, but also to
overfitted models with little practical relevance.

In the present study, automatic read-across like models were built to generate
quantitative predictions of long-term toxicity. The aim of the work was not to
predict the nature of the toxicological effects of chemicals, but to obtain
quantitative values which could be compared to exposure. Two databases
compiling chronic oral rat Lowest Adverse Effect Levels (LOAEL) as endpoint
were used. An early review of the databases revealed that many chemicals had at
least two independent studies/LOAELs. These studies were exploited to generate
information on the reproducibility of chronic animal studies and were used to
evaluate prediction performance of the models in the context of experimental
variability.

An important limitation often raised for computational toxicology is the lack
of transparency on published models and consequently on the difficulty for the
scientific community to reproduce and apply them. To overcome these issues,
source code for all programs and libraries and the data that have been used to
generate this manuscript are made available under GPL3 licenses. Data and
compiled programs with all dependencies for the reproduction of results in this
manuscript are available as a self-contained docker image. All data, tables and
figures in this manuscript was generated directly from experimental results
using the `R` package `knitR`.

Materials and Methods
=====================

The following sections give a high level overview about algorithms and datasets
used for this study. In order to provide unambiguous references to algorithms
and datasets, links to source code and data sources are included in the text.

Datasets
--------

### Nestlé database

The first database (Nestlé database for further reference) originates
from the publication of [@mazzatorta08]. It contains chronic (> 180 days) lowest
observed effect levels (LOAEL) for rats (*Rattus norvegicus*) after oral
(gavage, diet, drinking water) administration.  The Nestlé database consists
of `r length(m$SMILES)` LOAEL values for `r length(unique(m$SMILES))` unique
chemical structures.
The Nestlé database can be obtained from the following GitHub links:

  - original data: [https://github.com/opentox/loael-paper/blob/revision/data/LOAEL_mg_corrected_smiles_mmol.csv](https://github.com/opentox/loael-paper/blob/revision/data/LOAEL_mg_corrected_smiles_mmol.csv)
  - unique smiles: [https://github.com/opentox/loael-paper/blob/revision/data/mazzatorta.csv](https://github.com/opentox/loael-paper/blob/revision/data/mazzatorta.csv)
  - -log10 transfomed LOAEL: [https://github.com/opentox/loael-paper/blob/revision/data/mazzatorta_log10.csv](https://github.com/opentox/loael-paper/blob/revision/data/mazzatorta_log10.csv).

### Swiss Food Safety and Veterinary Office (FSVO) database

Publicly available data from pesticide evaluations of chronic rat
toxicity studies from the European Food Safety Authority (EFSA) [@EFSA2014], the Joint FAO/WHO Meeting on Pesticide Residues (JMPR) [@WHO2011] and the US EPA [@EPA2011] were compiled to form the
FSVO-database. Only studies providing both an experimental NOAEL and an
experimental LOAEL were included. The LOAELs were taken as they were
reported in the evaluations. Further details on the database are
described elsewhere [@Zarn2011, @Zarn2013]. The
FSVO-database consists of `r length(s$SMILES)` rat LOAEL values for `r length(unique(s$SMILES))` unique chemical
structures. It can be obtained from the following GitHub links:

  - original data: [https://github.com/opentox/loael-paper/blob/revision/data/NOAEL-LOAEL_SMILES_rat_chron.csv](https://github.com/opentox/loael-paper/blob/revision/data/NOAEL-LOAEL_SMILES_rat_chron.csv)
  - unique smiles and mmol/kg_bw/day units: [https://github.com/opentox/loael-paper/blob/revision/data/swiss.csv](https://github.com/opentox/loael-paper/blob/revision/data/swiss.csv)
  - -log10 transfomed LOAEL: [https://github.com/opentox/loael-paper/blob/revision/data/swiss_log10.csv](https://github.com/opentox/loael-paper/blob/revision/data/swiss_log10.csv)


### Preprocessing

Chemical structures (represented as SMILES [@doi:10.1021/ci00057a005]) in both
databases were checked for correctness. When syntactically incorrect or missing
SMILES were generated from other identifiers (e.g names, CAS numbers). Unique
smiles from the OpenBabel library [@OBoyle2011] were used for the
identification of duplicated structures. 

Studies with undefined or empty LOAEL entries were removed from the databases.
LOAEL values were converted to mmol/kg bw/day units and rounded to five
significant digits. For prediction, validation and visualisation purposes
-log10 transformations are used.

### Derived datasets

Two derived datasets were obtained from the original databases:

The [*test*
dataset](https://github.com/opentox/loael-paper/blob/revision/data/test_log10.csv)
contains data from compounds that occur in both databases. LOAEL values equal
at five significant digits were considered as duplicates originating from the
same study/publication and only one instance was kept in the test dataset. The
test dataset has `r length(t$SMILES)` LOAEL values for `r length(unique(t$SMILES))`
unique chemical structures and was used for

- evaluating experimental variability
- comparing model predictions with experimental variability.

The [*training*
dataset](https://github.com/opentox/loael-paper/blob/revision/data/training_log10.csv)
is the union of the Nestlé and the FSVO databases and it was used to build
predictive models. LOAEL duplicates were removed using the same criteria as for
the test dataset.  The training dataset has `r length(c$SMILES)` LOAEL values
for `r length(unique(c$SMILES))` unique chemical structures.

Algorithms
----------

In this study we are using the modular lazar (*la*zy *s*tructure *a*ctivity
*r*elationships) framework [@Maunz2013] for model development and validation.
The complete `lazar` source code can be found on [GitHub](https://github.com/opentox/lazar).

lazar follows the following basic [workflow](https://github.com/opentox/lazar/blob/loael-paper.revision/lib/model.rb#L180-L257):

For a given chemical structure lazar 

- searches in a database for similar structures (*neighbors*) with experimental
  data, 
- builds a local QSAR model with these neighbors and 
- uses this model to predict the unknown activity of the query compound.

This procedure resembles an automated version of *read across* predictions in
toxicology, in machine learning terms it would be classified as
a *k-nearest-neighbor* algorithm.

Apart from this basic workflow lazar is completely modular and allows the
researcher to use any algorithm for similarity searches and local QSAR
modelling. Algorithms used within this study are described in the following sections.

### Neighbor identification

Similarity calculations are based on [MolPrint2D fingerprints](https://github.com/opentox/lazar/blob/loael-paper.revision/lib/nanoparticle.rb#L17-L21)
[@doi:10.1021/ci034207y] from the OpenBabel chemoinformatics library
[@OBoyle2011].

The MolPrint2D fingerprint uses atom environments as molecular representation,
which resemble basically the chemical concept of functional groups. For each
atom in a molecule it represents the chemical environment using the atom types
of connected atoms.

MolPrint2D fingerprints are generated dynamically from chemical structures and
do not rely on predefined lists of fragments (such as OpenBabel FP3, FP4 or
MACCs fingerprints or lists of toxocophores/toxicophobes). This has the
advantage that they may capture substructures of toxicological relevance that
are not included in other fingerprints. 

From MolPrint2D fingerprints we can construct a feature vector with all atom
environments of a compound, which can be used to calculate chemical
similarities.

[//]: # https://openbabel.org/docs/dev/FileFormats/MolPrint2D_format.html#molprint2d-format

The [chemical similarity](https://github.com/opentox/lazar/blob/loael-paper.revision/lib/similarity.rb#L18-L20) between two compounds A and B is expressed as the
proportion between atom environments common in both structures $A \cap B$ and the
total number of atom environments $A \cup B$ (Jaccard/Tanimoto index, [@eq:jaccard]).

$$ sim = \frac{|A \cap B|}{|A \cup B|} $$ {#eq:jaccard}

The threshold selection is a trade-off between prediction accuracy (high
threshold) and the number of predictable compounds (low threshold). As it is in
many practical cases desirable to make predictions even in the absence of
closely related neighbors, we follow a tiered approach: 

- First a similarity threshold of 0.5 is used to collect neighbors, to create
  a local QSAR model and to make a prediction for the query compound.
- If any of these steps fails, the procedure is repeated with a similarity
  threshold of 0.2 and the prediction is flagged with a warning that it might
  be out of the applicability domain of the training data.
- Similarity thresholds of 0.5 and 0.2 are the default values chosen by the software developers and remained unchanged during the course of these experiments.

Compounds with the same structure as the query structure are automatically
[eliminated from neighbors](https://github.com/opentox/lazar/blob/loael-paper.revision/lib/model.rb#L180-L257)
to obtain unbiased predictions in the presence of
duplicates.

### Local QSAR models and predictions

Only similar compounds (*neighbors*) above the threshold are used for local
QSAR models.  In this investigation we are using [weighted random forests
regression
(RF)](https://github.com/opentox/lazar/blob/loael-paper.revision/lib/caret.rb#L7-L78)
for the prediction of quantitative properties.  First all uninformative
fingerprints (i.e. features with identical values across all neighbors) are
removed.  The remaining set of features is used as descriptors for creating
a local weighted RF model with atom environments as descriptors and model
similarities as weights. The RF method from the `caret` R package [@Kuhn08] is
used for this purpose.  Models are trained with the default `caret` settings,
optimizing the number of RF components by bootstrap resampling.

Finally the local RF model is applied to [predict the
activity](https://github.com/opentox/lazar/blob/loael-paper.revision/lib/model.rb#L194-L272)
of the query compound. The root-mean-square error (RMSE) of bootstrapped local model predictions is used
to construct 95\% prediction intervals at 1.96*RMSE. The width of the prediction interval indicates the expected prediction accuracy. The "true" value of a prediction should be with 95\% probability within the prediction interval.

If RF modelling or prediction fails, the program resorts to using the [weighted
mean](https://github.com/opentox/lazar/blob/loael-paper.revision/lib/regression.rb#L6-L16)
of the neighbors LOAEL values, where the contribution of each neighbor is
weighted by its similarity to the query compound. In this case the prediction
is also flagged with a warning.

### Applicability domain

The applicability domain (AD) of lazar models is determined by the structural
diversity of the training data. If no similar compounds are found in the
training data no predictions will be generated. Warnings are issued if the
similarity threshold has to be lowered from 0.5 to 0.2 in order to enable
predictions and if lazar has to resort to weighted average predictions, because
local random forests fail. Thus predictions without warnings can be considered
as close to the applicability domain and predictions with warnings as more
distant from the applicability domain. Quantitative applicability domain
information can be obtained from the similarities of individual neighbors.

Local regression models consider neighbor similarities to the query compound,
by weighting the contribution of each neighbor is by similarity. The
variability of local model predictions is reflected in the 95\% prediction
interval associated with each prediction.

### Validation

For the comparison of experimental variability with predictive accuracies we
are using a test set of compounds that occur in both databases. Unbiased read
across predictions are obtained from the *training* dataset, by [removing *all*
information](https://github.com/opentox/lazar/blob/loael-paper.revision/lib/model.rb#L234-L238)
from the test compound from the training set prior to predictions. This
procedure is hardcoded into the prediction algorithm in order to prevent
validation errors. As we have only a single test set no model or parameter
optimisations were performed in order to avoid overfitting a single dataset.

Results from 3 repeated [10-fold
crossvalidations](https://github.com/opentox/lazar/blob/loael-paper.revision/lib/crossvalidation.rb#L85-L93)
with independent training/test set splits are provided as additional
information to the test set results.

The final model for production purposes was trained with all available LOAEL
data (Nestlé and FSVO databases combined).

## Availability

Public webinterface
  ~ <https://lazar.in-silico.ch> (see [@fig:screenshot])

`lazar` framework
  ~ <https://github.com/opentox/lazar> (source code)

`lazar` GUI
  ~ <https://github.com/opentox/lazar-gui> (source code)

Manuscript
  ~ <https://github.com/opentox/loael-paper> (source code for the manuscript and validation experiments)

Docker image
  ~ <https://hub.docker.com/r/insilicotox/loael-paper/> (container with manuscript, validation experiments, `lazar` libraries and third party dependencies)

![Screenshot of a lazar prediction from the public webinterface.](figures/lazar-screenshot.pdf){#fig:screenshot}

Results
=======

### Dataset comparison

The main objective of this section is to compare the content of both
databases in terms of structural composition and LOAEL values, to
estimate the experimental variability of LOAEL values and to establish a
baseline for evaluating prediction performance.

##### Structural diversity

```{r echo=F}
fg = read.csv('data/functional-groups.csv',head=F)
```

In order to compare the structural diversity of both databases we evaluated the
frequency of functional groups from the OpenBabel FP4 fingerprint. [@fig:fg]
shows the frequency of functional groups in both databases. `r length(fg$V1)`
functional groups with a frequency > 25 are depicted, the complete table for
all functional groups can be found in the supplemental
material at [GitHub](https://github.com/opentox/loael-paper/blob/revision/data/functional-groups.csv).
 
![Frequency of functional groups.](figures/functional-groups.pdf){#fig:fg}

This result was confirmed with a visual inspection using the 
[CheS-Mapper](http://ches-mapper.org)  (Chemical Space Mapping and
Visualization in 3D, @Guetlein2012)
tool. 
CheS-Mapper can be used to analyze the relationship between the structure of
chemical compounds, their physico-chemical properties, and biological or toxic
effects. It depicts closely related (similar) compounds in 3D space and can be
used with different kinds of features. We have investigated structural as well
as physico-chemical properties and concluded that both databases are very
similar, both in terms of chemical structures and physico-chemical properties. 

The only statistically significant difference between both databases is that
the Nestlé database contains more small compounds (61 structures with less than
11 non-hydrogen atoms) than the FSVO-database (19 small structures, chi-square test: p-value 3.7E-7).

### Experimental variability versus prediction uncertainty 

Duplicated LOAEL values can be found in both databases and there is
a substantial number of `r length(unique(t$SMILES))` compounds with more than
one LOAEL. These chemicals allow us to estimate the variability of
experimental results within individual databases and between databases. Data with
*identical* values (at five significant digits) in both databases were excluded
from variability analysis, because it it likely that they originate from the
same experiments.

##### Intra database variability

```{r echo=F}
m.dupsmi <- unique(m$SMILES[duplicated(m$SMILES)])
s.dupsmi <- unique(s$SMILES[duplicated(s$SMILES)])
c.dupsmi <- unique(c$SMILES[duplicated(c$SMILES)])

m.dup <- m[m$SMILES %in% m.dupsmi,]
s.dup <- s[s$SMILES %in% s.dupsmi,]
c.dup <- c[c$SMILES %in% c.dupsmi,]

m.dupnr <- length(m.dupsmi)
s.dupnr <- length(s.dupsmi)
c.dupnr <- length(c.dupsmi)

m.dup$sd <- ave(m.dup$LOAEL,m.dup$SMILES,FUN=sd)
s.dup$sd <- ave(s.dup$LOAEL,s.dup$SMILES,FUN=sd)
c.dup$sd <- ave(c.dup$LOAEL,c.dup$SMILES,FUN=sd)
t$sd <- ave(t$LOAEL,t$SMILES,FUN=sd)

p = t.test(m.dup$sd,s.dup$sd)$p.value

m.mg = read.csv("data/mazzatorta_mg_dup.csv",header=T)
m.mg$sd <- ave(m.mg$LOAEL,m.mg$SMILES,FUN=sd)

s.mg = read.csv("data/swiss_mg_dup.csv",header=T)
s.mg$sd <- ave(s.mg$LOAEL,s.mg$SMILES,FUN=sd)

c.mg = read.csv("data/all_mg_dup.csv",header=T)
c.mg$sd <- ave(c.mg$LOAEL,c.mg$SMILES,FUN=sd)
```

Both databases contain substances with multiple measurements, which allow the determination of experimental variabilities. 
For this purpose we have calculated the mean LOAEL standard deviation of compounds with multiple measurements. Mean standard deviations and thus experimental variabilities are similar for both databases. 

The Nestlé database has `r length(m$SMILES)` LOAEL values for
`r length(levels(m$SMILES))` unique structures, `r m.dupnr` compounds have
multiple measurements with a mean standard deviation (-log10 transformed
values) of `r round(mean(m.dup$sd),2)` (`r round(mean(10^(-1*m.mg$sd)),2)`
mg/kg_bw/day, `r round(mean(10^(-1*m.dup$sd)),2)` mmol/kg_bw/day)
(@mazzatorta08, [@fig:intra]). 

The FSVO database has `r length(s$SMILES)` rat LOAEL values for
`r length(levels(s$SMILES))` unique structures, `r s.dupnr` compounds have
multiple measurements with a mean standard deviation (-log10 transformed values) of
`r round(mean(s.dup$sd),2)`
(`r round(mean(10^(-1*s.mg$sd)),2)` mg/kg_bw/day,
`r round(mean(10^(-1*s.dup$sd)),2)` mmol/kg_bw/day)
([@fig:intra]). 

Standard deviations of both databases do not show
a statistically significant difference with a p-value (t-test) of `r round(p,2)`.
The combined test set has a mean standard deviation (-log10 transformed values) of
`r round(mean(c.dup$sd),2)`
(`r round(mean(10^(-1*c.mg$sd)),2)` mg/kg_bw/day,
`r round(mean(10^(-1*c.dup$sd)),2)` mmol/kg_bw/day)
([@fig:intra]). 

![LOAEL distribution and variability of compounds with multiple measurements in both databases. Compounds were sorted according to LOAEL values. Each vertical line represents a compound, and each dot an individual LOAEL value. Experimental variability can be inferred from dots (LOAELs) on the same line (compound).](figures/dataset-variability.pdf){#fig:intra}

##### Inter database variability

In order to compare the correlation of LOAEL values in both databases and to establish a reference for predicted values, we have investigated compounds, that occur in both databases.

```{r echo=F}
data <- read.csv("data/median-correlation.csv",header=T)
cor <- cor.test(data$mazzatorta,data$swiss)
median.p <- cor$p.value
median.r.square <- round(rsquare(data$mazzatorta,data$swiss),2)
median.rmse <- round(rmse(data$mazzatorta,data$swiss),2)
``` 

[@fig:datacorr] depicts the correlation between LOAEL values from both
databases. As both databases contain duplicates medians were used for the
correlation plot and statistics. It should be kept in mind that the aggregation of duplicated
measurements into a single median value hides a substantial portion of the
experimental variability.  Correlation analysis shows a significant (p-value < 2.2e-16)
correlation between the experimental data in both databases with r\^2:
`r round(median.r.square,2)`, RMSE: `r round(median.rmse,2)`

[@fig:comp] shows the experimental LOAEL variability of compounds occurring in
both datasets (i.e. the *test* dataset) colored in blue (experimental). This is
the baseline reference for the comparison with predicted values.

![Correlation of median LOAEL values from Nestlé and FSVO databases. Data with
  identical values in both databases was removed from
  analysis.](figures/median-correlation.pdf){#fig:datacorr}

### Local QSAR models

```{r echo=F}
training = read.csv("data/training-test-predictions.csv",header=T)
warnings = subset(training,Warnings)
nowarnings = subset(training,!Warnings)
training.r_square = round(rsquare(training$LOAEL_measured_median,training$LOAEL_predicted),2)
training.rmse = round(rmse(training$LOAEL_measured_median,training$LOAEL_predicted),2)
warnings.r_square = round(rsquare(warnings$LOAEL_measured_median,warnings$LOAEL_predicted),2)
warnings.rmse = round(rmse(warnings$LOAEL_measured_median,warnings$LOAEL_predicted),2)
nowarnings.r_square = round(rsquare(nowarnings$LOAEL_measured_median,nowarnings$LOAEL_predicted),2)
nowarnings.rmse = round(rmse(nowarnings$LOAEL_measured_median,nowarnings$LOAEL_predicted),2)
misclassifications = read.csv("data/misclassifications.csv",header=T)
incorrect_predictions = length(misclassifications$SMILES)
correct_predictions = length(training$SMILES)-incorrect_predictions
```

In order to compare the performance of *in silico* read across models with
experimental variability we used compounds with multiple measurements as
a test set (`r  length(t$SMILES)` measurements, `r  length(unique(t$SMILES))`
compounds). `lazar` read across predictions were obtained for
`r length(unique(t$SMILES))` compounds, `r  length(unique(t$SMILES)) - length(training$SMILES)`
predictions failed, because no similar compounds were found in the training
data (i.e. they were not covered by the applicability domain of the training
data).

In `r round(100*correct_predictions/length(training$SMILES))`\% of the test examples
experimental LOAEL values were located within the 95\% prediction intervals. 

[@fig:comp] shows a comparison of predicted with experimental values. Most
predicted values were located within the experimental variability.

![Comparison of experimental with predicted LOAEL values. Each vertical line
represents a compound, dots are individual measurements (blue), predictions
(green) or predictions far from the applicability domain, i.e. with warnings
(red).](figures/test-prediction.pdf){#fig:comp}

Correlation analysis was performed between individual predictions and the
median of experimental data.  All correlations are statistically highly
significant with a p-value < 2.2e-16.  These results are presented in
[@fig:corr] and [@tbl:cv]. Please bear in mind that the aggregation of
multiple measurements into a single median value hides experimental variability.

Comparison    | $r^2$                     | RMSE    |  Nr. predicted
--------------|---------------------------|---------|---------------
Nestlé vs. FSVO database | `r median.r.square`      | `r median.rmse`           
AD close predictions vs. test median             | `r nowarnings.r_square` | `r nowarnings.rmse` | `r length(nowarnings$LOAEL_predicted)`/`r  length(unique(t$SMILES))`
AD distant predictions vs. test median             | `r warnings.r_square` | `r warnings.rmse`  | `r length(warnings$LOAEL_predicted)`/`r  length(unique(t$SMILES))`
All predictions vs. test median             | `r training.r_square` | `r training.rmse`  | `r length(training$LOAEL_predicted)`/`r  length(unique(t$SMILES))`

: Comparison of model predictions with experimental variability. {#tbl:common-pred}

![Correlation of experimental with predicted LOAEL values (test set). Green
dots indicate predictions close to the applicability domain (i.e. without
warnings), red dots indicate predictions far from the applicability domain
(i.e. with warnings).](figures/prediction-test-correlation.pdf){#fig:corr}

For a further assessment of model performance three independent 10-fold
cross-validations were performed. Results are summarised in [@tbl:cv] and
[@fig:cv]. All correlations of predicted with experimental values are
statistically highly significant with a p-value < 2.2e-16. This was observed for
compounds close and more distant to the applicability domain.

```{r echo=F}
cv50 = read.csv("data/50cv.csv",header=T)
```

Predictions  | $r^2$    | RMSE     |Nr. predicted 
-------------|----------|----------|----------------
AD close     | `r round(cv50$Mean[2],2)` $\pm$  `r round(cv50$SD[2],2)` | `r round(cv50$Mean[1],2)` $\pm$  `r round(cv50$SD[1],2)` | `r round(cv50$Mean[3],0)` $\pm$  `r round(cv50$SD[3],0)` 
AD distant   | `r round(cv50$Mean[5],2)`  $\pm$ `r round(cv50$SD[5],2)` | `r round(cv50$Mean[4],2)` $\pm$  `r round(cv50$SD[4],2)` | `r round(cv50$Mean[6],0)` $\pm$  `r round(cv50$SD[6],0)` 
All          | `r round(cv50$Mean[8],2)` $\pm$  `r round(cv50$SD[8],2)` | `r round(cv50$Mean[7],2)` $\pm$  `r round(cv50$SD[7],2)` | `r round(cv50$Mean[9],0)` $\pm$  `r round(cv50$SD[9],0)` 

: Results (mean and standard deviation) from 50 independent 10-fold crossvalidations {#tbl:cv}

![ Correlation of predicted vs. measured values from a randomly selected crossvalidation with MP2D fingerprint descriptors and local random forest
models.
](figures/crossvalidation.pdf){#fig:cv}

Discussion
==========

It is currently acknowledged that there is a strong need for
toxicological information on the multiple thousands of chemicals to
which human may be exposed through food. These include for example many
chemicals in commerce, which could potentially find their way into food
[@Stanton2016, @Fowler2011], but also substances
migrating from food contact materials [@Grob2006], chemicals
generated over food processing [@Cotterill2008], environmental
contaminants as well as inherent plant toxicants [@Schilter2013].
For the vast majority of these chemicals, no toxicological data
is available and consequently insight on their potential health risks is
very difficult to obtain. It is recognized that testing all of them in
standard animal studies is neither feasible from a resource perspective
nor desirable because of ethical issues associated with animal
experimentation. In addition, for many of these chemicals, risk may be
very low and therefore testing may actually be irrelevant. In this
context, the identification of chemicals of most concern on which
limited resource available should focused is essential and computational
toxicology is thought to play an important role for that.

In order to establish the level of safety concern of food chemicals
toxicologically not characterized, a methodology mimicking the process of
chemical risk assessment, and supported by computational toxicology, was
proposed [@Schilter2014]. It is based on the calculation of margins of exposure
(MoE) that is the ratio between the predicted chronic toxicity value (LOAEL)
and exposure estimate. The level of safety concern of a chemical is then
determined by the size of the MoE and its suitability to cover the
uncertainties of the assessment. To be applicable, such an approach requires
quantitative predictions of toxicological endpoints relevant for risk
assessment. The present work focuses on the prediction of chronic toxicity,
a major and often pivotal endpoint of toxicological databases used for hazard
identification and characterization of food chemicals.

In a previous study, automated read-across like models for predicting
carcinogenic potency were developed. In these models, substances in the
training dataset similar to the query compounds are automatically
identified and used to derive a quantitative TD50 value. The errors
observed in these models were within the published estimation of
experimental variability [@LoPiparo2014]. In the present
study, a similar approach was applied to build models generating
quantitative predictions of long-term toxicity. Two databases compiling
chronic oral rat lowest adverse effect levels (LOAEL) as reference value were
available from different sources. Our investigations clearly indicated that the
Nestlé and FSVO databases are very similar in terms of chemical
structures and properties as well as distribution of experimental LOAEL
values. The only significant difference that we observed was that the
Nestlé one has larger amount of small molecules, than the FSVO database.
For this reason we pooled both databases into a single training dataset
for read across predictions.

An early review of the databases revealed that 155 out of the 671
chemicals available in the training datasets had at least two
independent studies/LOAELs. These studies were exploited to generate
information on the reproducibility of chronic animal studies and were
used to evaluate prediction performance of the models in the context of
experimental variability. Considerable variability in the experimental
data was observed. Study design differences, including dose selection,
dose spacing and route of administration are likely explanation of
experimental variability. High experimental variability has an impact on
model building and on model validation. First it influences model
quality by introducing noise into the training data, secondly it
influences accuracy estimates because predictions have to be compared
against noisy data where "true" experimental values are unknown. This
will become obvious in the next section, where comparison of predictions
with experimental data is discussed. The data obtained in the present study indicate
that `lazar` generates reliable predictions for compounds within the
applicability domain of the training data (i.e. predictions without
warnings, which indicates a sufficient number of neighbors with
similarity &gt; 0.5 to create local random forest models). Correlation
analysis shows that errors ($\text{RMSE}$) and explained variance
($r^{2}$) are comparable to experimental variability of the training
data.

Predictions with a warning (neighbor similarity &lt; 0.5 and &gt; 0.2 or
weighted average predictions) are more uncertain. However, they still
show a strong correlation with experimental data, but the errors are ~ 20-40\%
larger than for compounds within the applicability domain ([@fig:corr] and [@tbl:cv]). Expected
errors are displayed as 95% prediction intervals, which covers 100% of
the experimental data. The main advantage of lowering the similarity
threshold is that it allows to predict a much larger number of
substances than with more rigorous applicability domain criteria. As
each of this prediction could be problematic, they are flagged with a
warning to alert risk assessors that further inspection is required.
This can be done in the graphical interface
(<https://lazar.in-silico.ch>) which provides intuitive means of
inspecting the rationales and data used for read across predictions.

Finally there is a substantial number of chemicals (37), where no
predictions can be made, because no similar compounds in the training
data are available. These compounds clearly fall beyond the
applicability domain of the training dataset and in such cases
predictions should not be used. In order to expand the domain of
applicability, the possibility to design models based on shorter, less
than chonic studies should be studied. It is likely that more substances
reflecting a wider chemical domain may be available. To predict such
shorter duration endpoints would also be valuable for chronic toxicy
since evidence suggest that exposure duration has little impact on the
levels of NOAELs/LOAELs [@Zarn2011, @Zarn2013].

### `lazar` predictions

[@tbl:common-pred], [@tbl:cv], [@fig:comp], [@fig:corr] and [@fig:cv] clearly
indicate that `lazar` generates reliable predictions for compounds within the
applicability domain of the training data (i.e. predictions without warnings,
which indicates a sufficient number of neighbors with similarity > 0.5 to
create local random forest models). Correlation analysis ([@tbl:common-pred],
[@tbl:cv]) shows, that errors ($RMSE$) and explained variance ($r^2$) are
comparable to experimental variability of the training data.

Predictions with a warning (neighbor similarity < 0.5 and > 0.2 or weighted
average predictions) are a grey zone. They still show a strong correlation with
experimental data, but the errors are larger than for compounds within the
applicability domain ([@tbl:common-pred], [@tbl:cv]). Expected errors are
displayed as 95\% prediction intervals, which covers
`r round(100*correct_predictions/length(training$SMILES))`\% of the experimental
data. The main advantage of lowering the similarity threshold is that it allows
to predict a much larger number of substances than with more rigorous
applicability domain criteria. As each of this prediction could be problematic,
they are flagged with a warning to alert risk assessors that further inspection
is required. This can be done in the graphical interface
(<https://lazar.in-silico.ch>) which provides intuitive means of inspecting the
rationales and data used for read across predictions.

Finally there is a substantial number of compounds
(`r length(unique(t$SMILES))-length(training$LOAEL_predicted)`),
where no predictions can be made, because there are no similar compounds in the training data. These compounds clearly fall beyond the applicability domain of the training dataset 
 and in such cases it is preferable to avoid predictions instead of random guessing.

Summary
=======

In conclusion, we could demonstrate that `lazar` predictions within the
applicability domain of the training data have the same variability as the
experimental training data. In such cases experimental investigations can be
substituted with *in silico* predictions. Predictions with a lower similarity
threshold can still give usable results, but the errors to be expected are
higher and a manual inspection of prediction results is highly recommended.
Anyway, our suggested workflow includes always the visual inspection of the
chemical structures of the neighbors selected by the model. Indeed it will
strength the prediction confidence (if the input structure looks very similar
to the neighbors selected to build the model) or it can drive to the conclusion
to use read-across with the most similar compound of the database (in case not
enough similar compounds to build the model are present in the database).

References
==========
